---
output:
  html_document: default
  pdf_document: default
---
#GIVE ME MORE WINE 
##by NISHCHAY CHAWLA


========================================================



```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

library(ggplot2)
library(gridExtra)
library(GGally)

```

```{r, echo=FALSE}
whw <- read.csv('wineQualityWhites.csv')
#getting started
p1 <- ggplot(data = whw) + theme_dark()
whw$X <- NULL
whw$quality <-  ordered(whw$quality)
```




## Introduction

The dataset is related to white variant of the Portuguese "Vinho Verde" wine. 
The dataset points out few of many factors which inflence world's all time 
favourite beverage (WINE!!).Type of grapes, yeast strain used, age, temperature,
aging process, are generally the terms understood by enthusiasts. Although the 
dataset include variables based on physicochemical tests only, 
it will be very intersting to draw patterns in the technicalities of wines 
judged by the experts.


### Citation
P. Cortez, A. Cerdeira, F. Almeida, T. Matos and J. Reis. 
  Modeling wine preferences by data mining from physicochemical properties.
  In Decision Support Systems, Elsevier, 47(4):547-553. ISSN: 0167-9236.

  Available at: [@Elsevier] http://dx.doi.org/10.1016/j.dss.2009.05.016
                [Pre-press (pdf)] http://www3.dsi.uminho.pt/pcortez/winequality09.pdf
                [bib] http://www3.dsi.uminho.pt/pcortez/dss09.bib


# Univariate Plots Section

### Structure Of Data:


```{r, echo=FALSE}
str(whw)

```

The orignal dataset has quality as numeric variable but it is converted to an 
ordered variable for the sake of exploration.

### Summary Of Data


```{r, echo=FALSE}
summary(whw)
```



```{r,echo=FALSE, warning=FALSE, message=FALSE}
# Following function will be used for creating histogram matrix as well as 
#individual histogram
# Plotting some variables together will make visualisation and explination easier.

# Binwidths are slected by trails, the one which made histogram visually good and 
# made explination easy.

his.matrix <- function(abc, bin) {
    binwidth <- bin 
     ggplot(data = whw, aes_string(x = abc )) + 
    geom_histogram( fill ='#44D7A8', colour = '#FFFF99', binwidth = binwidth) +
    geom_vline(aes(xintercept = mean(whw[,abc] )) , color = 'purple', 
               linetype = 'longdash', size = .75 ) +
    geom_vline(aes(xintercept = median(whw[,abc] )), color = 'orange',
                size = .8) +
    geom_vline(aes(xintercept = quantile(whw[,abc], 0.25 )), 
               linetype = 'dotted', size=0.5) + 
    geom_vline( aes(xintercept=quantile(whw[,abc], 0.75 )), 
               linetype = 'dotted', size=0.5)+
    theme_dark() } 
```


```{r echo=FALSE, Univariate_Plots, warning=FALSE, message=FALSE}
#grid of histograms of acid variables using his.matrix function and binwidth
# of .009

variable_a <- c("volatile.acidity","citric.acid")
plots_a <- lapply(variable_a, his.matrix, bin=.009) 

do.call(grid.arrange, args = c(plots_a, ncol = 1, top = 'Acidity'))



```




> *Purple-Dashed line indicates the mean, orange indicates the median,\
while the dotted lines are for quartiles.*
*1 dm^3 = 1 Liter*

Both Citric and Acetic acid(indicated by volatile acid) are measured in gm/dm^3
and have an almost normal distribution. Citric acid have an unusual peak at 
0.49 gm/dm^3, this may be because of persence of many wines from one paticular 
winemaker or because of any regulations.

```{r, echo=FALSE,  warning=FALSE, message=FALSE}
#grid of histograms of acid variable using his.matrix function and binwidth
# of .09
variable_f <- c("fixed.acidity")
plots_f <- lapply(variable_f, his.matrix, bin=.09) 

do.call(grid.arrange, args = c(plots_f, ncol = 1))
```



Fixed acidity indicates tartaric acid measured in gm/dm^3 is also almost normally
distributed.




```{r, echo=FALSE, warning=FALSE, message=FALSE}
#grid of histograms of chloride using his.matrix function and binwidth
# of .003

variable_d <- c("chlorides" )
plots_d <- lapply(variable_d, his.matrix, bin = .003)

do.call(grid.arrange, args = c(plots_d, ncol = 1, top = 'Chloride'))


```




The chloride data looks very lepokurtic as most data lies in 0.1 range.
Infact 3rd Quartile at 0.05 indicates 75% data lies between 0.009 an 0.05
(0.009 being min. value). Although data goes upto .34, very very few value lies
above .1.


```{r,echo=FALSE,  warning=FALSE, message=FALSE}
#grid of sulphur variables
variable_b <- c("free.sulfur.dioxide","total.sulfur.dioxide" )
plots_b <- lapply(variable_b, his.matrix, bin = 2)

do.call(grid.arrange, args = c(plots_b, ncol = 1, top = 'Sulphur Dioxide'))
```



Both Free and Total Sulphur dioxied have symmetric data and both are measured in
mg/dm^3


```{r, echo=FALSE,  warning=FALSE, message=FALSE}

#grid of histograms of density variables using his.matrix function and binwidth
# of .0006
variable_e <- c("density" )
plots_e <- lapply(variable_e, his.matrix, bin = .0006)
do.call(grid.arrange, args = c(plots_e, ncol = 1, top = 'Density'))



```






The density data is present in extremely narrow range. As indicated from summary
and the visualisation above min value is .9871 and 3 Qt is at .9961. This is an
very interesting observation because with almost 4900 observations this small range 
can help making a generalised statement like wines have density of .99 gm/cm^3.
This value can be used for doing mathematical calculations.



```{r, echo=FALSE, warning=FALSE, message=FALSE}

#grid of histograms of pH using his.matrix function and binwidth
# of .009

variable_h <- c("pH")
plots_h <- lapply(variable_h, his.matrix, bin = .009)

do.call(grid.arrange, args = c(plots_h, ncol = 1))
```



pH range of wine in our datset 2.7 and 3.8, with average of 3.18. 
pH is basically a measure of acidity with 0 being most acidic , 7, neutral and 
14 being highly basic. 


```{r, echo=FALSE, warning=FALSE, message=FALSE}

#grid of histograms of sulphate variables using his.matrix function and binwidth
# of .01
variable_c <- c("sulphates")
plots_c <- lapply(variable_c, his.matrix, bin = .01)

do.call(grid.arrange, args = c(plots_c, ncol = 1))


```




Sulphates are also measured in gm/dm^3 and the data have some peaks.

```{r, echo=FALSE,  warning=FALSE, message=FALSE}

#grid of histograms of alcohol variables using his.matrix function and binwidth
# of .07
variable_i <- c("alcohol")
plots_i <- lapply(variable_i, his.matrix, bin = .07)

do.call(grid.arrange, args = c(plots_i, ncol = 1))


```




Alcohol measured in % volume dosen't show much symmetry and have values between 
8 and 14 with 75% lying below 11.4. As visible we have most no. of wines 
with alcohol percent between 9.5 and 9.6.




```{r, echo=FALSE, message=FALSE}

p10 <- p1+ aes(x = residual.sugar)+
  geom_histogram(fill =I('#44D7A8'), binwidth = .009 )+
  scale_x_log10(breaks = seq(1,15,1))+
  geom_vline(aes(xintercept = median(whw$residual.sugar )), color = 'pink',
             size = .8) +
  geom_vline(aes(xintercept = mean(whw$residual.sugar)) , color = 'purple', 
             linetype = 'longdash', size = .75 )+
  geom_vline(aes(xintercept = quantile(whw$residual.sugar, 0.25 )), 
             linetype = 'dotted', size=0.5) + 
  geom_vline( aes(xintercept=quantile(whw$residual.sugar, 0.75 )), 
              linetype = 'dotted', size=0.5)+
  xlab('residual sugar (Log scale)')


variable_g <- c("residual.sugar" )
plots_g <- lapply(variable_g, his.matrix, bin = .5)

plots_g[[2]] <- p10
do.call(grid.arrange, args = c(plots_g, ncol = 1, top = "Resdidual Sugar"))



```



* Purple-Dashed line indicates the mean, orange indicates the median, 
while the dotted lines are for quartiles.


Residual Sugar, or RS for short, refers to any natural grape sugars that 
is leftover after fermentation ceases. When plotted on log scale showed 
much better visualisation. Half of the wines have RS between 0.5 and
5 gm/liter while the other half lies between 5 and 20 gm/liter. The data seems to
have a normal distribution betweem 0 and 5 gm/liter on log scale.

```{r,echo=FALSE, message=FALSE}
p10 <- p1+ aes(x = quality)+
  geom_bar(fill ='#44D7A8',colour = '#FFFF99')
p10
```


Quality :  score between 0 and 10
```{r,echo=FALSE, message=FALSE}

summary(whw$quality)
```

Most wines were rated 5 or 6, only 5 were rated 9



# Univariate Analysis


### What is the structure of your dataset?

The data has **11 parameters** and expert rating of **4898 wines**.

*Input variables* :

1. fixed acidity (tartaric acid - g / dm^3) 

2. volatile acidity (acetic acid - g / dm^3) 

3. citric acid (g / dm^3) 

4. residual sugar (g / dm^3) 

5. chlorides (sodium chloride - g / dm^3 

6. free sulfur dioxide (mg / dm^3) 

7. total sulfur dioxide (mg / dm^3) 

8. density (g / cm^3) 

9. pH 

10. sulphates (potassium sulphate - g / dm3) 

11. alcohol (% by volume)  

*Output variable*:
1. Quality (score between 0 and 10) # Missing Attribute Values: None

### What is/are the main feature(s) of interest in your dataset?

1. Most of the wine are rated 5 or 6 by the experts, this might be an intersting
point because generally human judgement tend to follow a normal distribution, 
i.e. most of us stays in between and avoid extremities. 
Although experts, judge on strict parameters but when close to 4900 wine are 
judged, a normal-ish pattern emerged.
2. Yeast feasts on sugar to produce alcohol in fermentation process, 
also alcohol and residual sugar dont follow the general trend of normal 
distribution, this may reveal some new patterns.


### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

1. pH values conveys basic detials about acidity and acidity inturn 
plays an important role in providing wine its taste, so it might be an 
intersting factor.
2. Production of alcohol also varies densities.
3. Sulfur dioxide is an integral part of wine making, so it can't be ignored here.

### Did you create any new variables from existing variables in the dataset?
No new variables were created

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?
1. When observed closely on log scale residual sugar seemed to follow normal
distribution in intervals. Log scale was chosen because the data showed 
skewness and most of the values were near 1 and 10
2. Quality variable was present as numeric variable initially, 
it was converted to ordered factor to make it more like a ranking system.

***

# Bivariate Plots Section

```{r ,echo=FALSE, Bivariate_Plots, message=FALSE, warning=FALSE,fig.width=9, fig.height= 8 }

ggpairs(data = whw,
        lower = list(continuous = wrap("points", shape = I('.'))),
        upper = list(combo = wrap("box", outlier.shape = I('.'))))


```



The upper triangle show corelation and lower triangle has the scatterplots 
between all the quantitative variables of dataset.
Such matrix makes it very easy to pinpoint variables with noticeable pattern

```{r, echo=FALSE, message=FALSE, warning= FALSE}

p11 <- ggplot(data= whw, aes(y = alcohol,x = quality))+
  geom_boxplot(colour = "#44D7A8")+
  geom_jitter(alpha=1/20)+
  theme_dark()+
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "red", 
               shape = 8, 
               size = 4)
p11

```



A box plot is an efficient way of visualising continous data against 
categorical data. Upper line of box indicates 3qt, 
lower indicates 1Qt while solid line inside the box indicates the median.
A trend is clearly visible in the above plot.

```{r , echo=FALSE, message=FALSE}
ggplot(data = whw,aes(y = chlorides,x = quality ))+
  geom_boxplot(colour = "#44D7A8")+
  geom_jitter(alpha=1/20)+
  theme_dark()+
  coord_cartesian(ylim = c(0,0.15))+
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "red", 
               shape = 8, 
               size = 4)
```


```{r, echo=FALSE, message=FALSE}
ggplot(data = whw, aes(x = quality, y = pH))+
  geom_boxplot(colour = "#44D7A8")+
  geom_jitter(alpha=1/20)+
  theme_dark()+
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "red", 
               shape = 8, 
               size = 4)
```



White wines of all ratings have pH in the range 3 to 3.3,  except for very few.
The " * " indicates the mean value of variables in different qualities.


```{r, echo=FALSE, message=FALSE, warning= FALSE}
p2 <- ggplot(data = whw, aes(x = quality, fill = quality))+
  geom_boxplot()+
  scale_fill_brewer(type = 'qual', palette = 6, guide = F)+
  theme_dark()

p3 <- p2+aes(y = total.sulfur.dioxide)+
  ylab('Total Sulphur Dioxide')
p4 <- p2+aes(y = free.sulfur.dioxide)+
  ylab('Free Sulphur Dioxoide' )
p5 <- p1+ aes(x =alcohol, y = total.sulfur.dioxide)+
  geom_point(colour='#44D7A8', alpha = 1/4)+
  geom_smooth(method = 'lm')+
  ylab('Total Sulphur Dioxide')
p6 <- p1+ aes(x =alcohol, y = free.sulfur.dioxide)+
  geom_point(colour='#44D7A8', alpha = 1/4)+
  geom_smooth(method = 'lm')+
  ylab('Free Sulphur Dioxoide')


grid.arrange(p3,p4,p5,p6, top = 'Sulphur Relations')
```



No paticular pattern emerged from the above graphs.

```{r, echo=FALSE, message=FALSE}

p1+aes(y = alcohol, x = residual.sugar)+
  geom_point(colour='#44D7A8', alpha = 1/4)+
  geom_smooth(method='lm')+
  coord_cartesian(ylim = c(7.5,14), xlim=c(0,23))+
  xlab('residual sugar (g / dm^3)')+
  ggtitle('Alcohol vs Residual Sugar')
  
```


```{r, echo=FALSE, message=FALSE}
p1+aes(x = density, y =alcohol)+
  geom_point(colour='#44D7A8', alpha = 1/4)+
  geom_smooth(method='lm')+
  coord_cartesian(xlim = c(0.985,1.0025), ylim = c(7.5,15))+
  ggtitle('Alcohol vs Density')
```


Alcohol have negative correlatiion with Residual sugar and density, 
and a similar trend is observed in the scatterplots.

```{r, echo=FALSE, message=FALSE}
p1+ aes(x= chlorides, y = alcohol)+
  geom_point(colour='#44D7A8', alpha = 1/3)+
  coord_cartesian(xlim= c(0,0.2), ylim=c(7.5,14))+
  geom_smooth(method='lm')+
  ggtitle('Alcohol vs Chlorides')
```



The majority of wines have chlorides between .01 and .075.

# Bivariate Analysis



### Talk about some of the relationships you observed in this part of the \n investigation. How did the feature(s) of interest vary with other features in \n the dataset?

Some negative correlations were observed in mathematical and visual form.
The negative correlation between alcohol and residual sugar made complete sense, 
because yeast consume sugar to produce alcohol, and the sugar which is left 
after the fermentation ceases is the residual sugar. Infact one of the reason for 
fermentation reaction to stop is the increase in percentage of alcohol, 
yeast generally don't survive once 12%-14% concerntration is achieved and this
is also visible in our data as wine having max alcohol has 14.2% .

Another interesting trend was wine rated higher have more alcohol percentage.


### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

Variation of density with alcohol percent. The density decreases as alcohol
increases.

### What was the strongest relationship you found?
1. The ranking of higher alcohol percentage wines was clearly more.
2. ALcohol and density had a strong negative corelation.

***

# Multivariate Plots Section


```{r echo=FALSE, Multivariate_Plots}
p12 <- ggplot(data = whw ,aes(x = residual.sugar, y = density, colour = quality ))+
  geom_point(alpha =1/2)+
  geom_smooth(method = "lm", se = F, size = .5)+
  scale_color_brewer(type = 'div', palette =  'RdBu')+ theme_dark()+
  coord_cartesian(xlim=c(0,22), ylim = c(0.985,1.0025))

p12

```



Low density and low residual sugar i.e bottom left part of graph sees 
overcrowding of higher rated wines.
The Lines are plotted according to a linear regression model.

```{r, echo=FALSE, message=FALSE}
ggplot(data = whw, aes(x = density, y = alcohol, colour = quality ))+
  geom_point(alpha = 1/2)+
  geom_smooth(method = "lm", se = F, size = .5)+
  scale_color_brewer(type = 'div', palette =  'RdBu')+ theme_dark()+
  coord_cartesian( xlim = c(0.985,1.0025), ylim = c(7,14))
```



The graph shows two clusters one with rating 7 and obove with density .990 and 
second one with rating 6 and below with density .995-1.00.


```{r, echo=FALSE, message=FALSE}
ggplot(data = whw, aes(x = residual.sugar, y = alcohol, colour = quality ))+
  geom_point(alpha =1/2)+
   geom_smooth(method = "lm", se = F, size = .5)+
  scale_color_brewer(type = 'div', palette =  'RdBu')+ theme_dark()+
  coord_cartesian( xlim = c(0,22), ylim = c(7,14))
```





The above three graphs shows that higher rated wine have density and residual
sugar on the lower side and alcohol on the higher as side, the Lines are 
plotted according to a linear regression model, which makes it even more clearer
to visualise that. 



# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

As observed earlier higher rated wines tend to have higher alcohol percentage 
which inturn is negatively corelated with residual sugar and density, 
so when all these variables were plotted together, the whole concept made more sense 


### Were there any interesting or surprising interactions between features?
The plotted variables interacted as expected.


------

# Final Plots and Summary


### Plot One
```{r echo=FALSE, Plot_One}

p10 +  xlab('Quality (score between 0 and 10)')+
  ggtitle('Quality Bar Chart')
```




>“Accept what life offers you and try to drink from every cup. 
All wines should be tasted; some should only be sipped, but with others, 
drink the whole bottle.” 
    ― Paulo Coelho, Brida




### Description One
This one simple graph is conveying a lot more than just ranking,
there is a high possibility it is defining human behaviour. 
A simple search on google about about wine making provides methods to make your own wine.
Apart from some basic framework the wine making process is very flexible, 
and we generally try to make thing just right. 
Not too spicy. Not too sweet. Just right. Most of the things we buy or make 
have to fit in that middle ground. The whole process from selecting grapes to 
fermentation to ageging is relying on the different definitions of just right, 
maybe this is the reason there are very few wines with rating 3 or 9. 
That is why the presence of normal distribution in most of the parameters 
makes a lot of sense.
Although, it goes without saying that what might be 3 for one will be a 
9 for other, but that also brings a lot of other factor in picture,
anyhow in general human behaviour follows a normal distribution. 




### Plot Two
```{r echo=FALSE, Plot_Two}

  p11 + ggtitle('Alcohol content in different Qualities')+
  ylab('alcohol (% by volume)')+
  xlab('Quality (score between 0 and 10)' ) 
```




### Description Two
The graph showed a significant pattern that more alcohol 
content returns better rating.
This might also be an insight to rating sysytem because ~1% change in median
alcohol value in wines with rating 5 and above, improved the rating by one unit. 




### Plot Three
```{r echo=FALSE, Plot_Three}
p12 + ggtitle('Density vs Residual Sugar', subtitle = 'As per quality')+
  ylab('density (g / cm^3)')+
  xlab('residual sugar (g / dm^3)' ) 
```






### Description Three
This graph shows all the key observations made among the parameters.
Higher rated alcohol has low sugar and low density. Fermentation ceases once
alcohol percentage is in between 12-14 (remember alcohol and residual sugar are
negatively co-related), thus reaching that levels is like getting the maximum 
out of yeast.
Fermentation also depend on other factors like temperature, acidity etc 
and mantaining those at higher level of alcohol(low residual sugar) 
can be even more challenging, thus managing to get those higher level is 
clearly rewarded by experts.
------




# Reflection


I started by studying some basic fundamentals about wine making to get 
a better understanding of data like the role of sulphur dioxide, acidity etc.
Then proceeded to make some random graphs matrix which helped me selcting some 
key variables like alcohol percentage, residual sugar. Working on this dataset 

The persence of normal distribution among most values facinated me alot but that
also became a challenge, because one can point out abnormalities and express
those in words. With normal distributions only few options like
mean, median, quartile range, come to mind initially but on close observations and
patience data started making sense.
Another challenge was making a clean code which was overcomed by use of custom
functions

In future I hope to create a model for predicting the quality of wine, also 
innclusion of price, age of the wine and adding more regions could make this
dataset even more exciting.

> “Either give me more wine or leave me alone.” 
 ― Jalaluddin Rumi
